---
title: "biostat203bhw4"
format: html
editor: visual
author: Yasmine Tani; UID 406579988
---

# Biostatistics 203B - Homework 4

```{r}
sessionInfo()
```

```{r}
memuse::Sys.meminfo()
```

```{r}
library(bigrquery)
library(dbplyr)
library(DBI)
library(gt)
library(gtsummary)
library(tidyverse)
```

## **Q1. Compile the ICU cohort in HW3 from the Google BigQuery database**

### **Q1.1 Connect to BigQuery**

```{r}
# path to the service account token
satoken <- "biostat-203b-2026-winter-92fefbfab477.json"
# BigQuery authentication using service account
bq_auth(path = satoken)
```

```{r}
# connect to the BigQuery database `biostat-203b-2025-mimiciv_3_1`
con_bq <- dbConnect(
  bigrquery::bigquery(),
  project = "biostat-203b-2025-winter",
  dataset = "mimiciv_3_1",
  billing = "biostat-203b-2025-winter"
)
con_bq
```

```{r}
dbListTables(con_bq)
```

### **Q1.2 ICU Stays Data**

```{r}
# full ICU stays table
icustays_tble <- tbl(con_bq, "icustays") |>
  arrange(subject_id, hadm_id, stay_id) |>
  # show_query() |>
  print(width = Inf)
```

### **Q1.3 Admissions Data**

```{r}
admissions_tble <- tbl(con_bq, "admissions") |>
  arrange(subject_id, hadm_id) |>
  # show_query() |>
  print(width = Inf)
```

### **Q1.4 Patients Data**

```{r}
patients_tble <- tbl(con_bq, "patients") |>
  arrange(subject_id) |>
  print(width = Inf)
```

### **Q1.5 LabEvents Data**

```{r}
labevents_tble <- tbl(con_bq, "labevents") %>%
  # hw3 labs that we want
  filter(itemid %in% c(50912, 50971, 50983, 50902, 50882, 51221, 51301, 50931)) %>%
  # keep only subjects in icustays_tble
  semi_join(icustays_tble, by = "subject_id") %>%
  # join with icustays to get the intime
  left_join(
    icustays_tble %>% select(subject_id, stay_id, intime),
    by = "subject_id"
  ) %>%
  # labs taken BEFORE ICU admission
  filter(storetime <= intime) %>%
  # very last measurement
  group_by(subject_id, stay_id, itemid) %>%
  slice_max(order_by = storetime, n = 1) %>%
  ungroup() %>%
  select(subject_id, stay_id, itemid, valuenum) %>%
  # change itemid codes into column names
  pivot_wider(names_from = itemid, values_from = valuenum) %>%
  # making IDs the atcul lab names
  rename(
    creatinine = `50912`,
    potassium = `50971`,
    sodium = `50983`,
    chloride = `50902`,
    bicarbonate = `50882`,
    hematocrit = `51221`,
    wbc = `51301`,
    glucose = `50931`
  ) %>%
  arrange(subject_id, stay_id) %>%
  collect()
```

```{r}
# antoinette advised we check space to ensure we did properly
nrow(labevents_tble)

head(labevents_tble)
```

### **Q1.6 chartevents data**

```{r}
chartevents_tble <- tbl(con_bq, "chartevents") |>
  filter(itemid %in% c(220045, 220179, 220180, 220210, 223761)) |>
  semi_join(icustays_tble, by = "subject_id") |>
  left_join(
    icustays_tble |> select(subject_id, stay_id, intime, outtime),
    by = c("subject_id", "stay_id")
  ) |>
  filter(charttime >= intime & charttime <= outtime) |>
  group_by(subject_id, stay_id, itemid) |>
  filter(charttime == min(charttime, na.rm = TRUE)) |>
  summarize(
    valuenum = mean(valuenum, na.rm = TRUE),
    .groups = "drop"
  ) |>
  pivot_wider(names_from = itemid, values_from = valuenum) |>
  rename(
    heart_rate  = `220045`,
    sys_bp      = `220179`,
    dias_bp     = `220180`,
    resp_rate   = `220210`,
    temp_f      = `223761`
  ) |>
  collect()
```

```{r}
nrow(chartevents_tble)
```

### **Q1.7 Putting Things Together**

```{r}
mimic_icu_cohort <- icustays_tble |>
  # merging
  left_join(admissions_tble, by = c("subject_id", "hadm_id")) |>
  left_join(patients_tble, by = "subject_id") |>
  collect() |>
  # adults only age calculation like b4 (tho i think boston is only adults but)
  mutate(age = year(intime) - (anchor_year - anchor_age)) |>
  filter(age >= 18) |>
  # merge w labevents and chartevents tables
  left_join(labevents_tble, by = c("subject_id", "stay_id")) |>
  left_join(chartevents_tble, by = c("subject_id", "stay_id")) |>
  arrange(subject_id, hadm_id, stay_id) |>
  print(width = Inf)
```

### **Q1.8 Preprocessing**

```{r}
library(forcats)
library(gtsummary)

mimic_icu_cohort <- mimic_icu_cohort |>
  mutate(across(
    c(
      first_careunit, last_careunit, admission_type,
      admission_location, discharge_location
    ),
    ~ fct_lump_n(.x, n = 4, other_level = "Other")
  )) |>
  mutate(race = fct_collapse(race,
    WHITE = grep("WHITE", levels(as.factor(race)), value = TRUE),
    BLACK = grep("BLACK", levels(as.factor(race)), value = TRUE),
    HISPANIC = grep("HISPANIC", levels(as.factor(race)), value = TRUE),
    ASIAN = grep("ASIAN", levels(as.factor(race)), value = TRUE),
    other_level = "Other"
  )) |>
  mutate(los_long = los >= 2)

# re-save the file
write_rds(mimic_icu_cohort, "mimiciv_shiny/mimic_icu_cohort.rds")

mimic_icu_cohort |>
  # variables the proff showed
  select(los_long, first_careunit, last_careunit, admission_type, race, age, gender, marital_status, language, age) |>
  # summary
  tbl_summary(by = los_long) |>
  add_p()
```

### 

### **Q1.9 Save the final tibble**

```{r}
# make a directory mimiciv_shiny
if (!dir.exists("mimiciv_shiny")) {
  dir.create("mimiciv_shiny")
}
# save the final tibble
if (exists("mimic_icu_cohort")) {
  mimic_icu_cohort |>
    write_rds("mimiciv_shiny/mimic_icu_cohort.rds", compress = "gz")
}
```

```{r}
if (exists("con_bq")) {
  dbDisconnect(con_bq)
}
rm(list = ls())
```

### AI Assistant

I used Gemini 3 Flash, usually on fast thinking but also set to thinking mode when I really needed it to 'focus.' I sometimes clarified that if it did suggest code, it should follow the tidyverse style guide though I now run r styler at the end, so perhaps that is not needed.

1.  I asked with help for my dropdowns in tab 1 to get the graphics to summarize the los, gender, etc. Instead, it displayed every single variable even noisy ones so I had a dropdown with 40+ variables that were not insightful for the user.
2.  Instead of ensuring I used json token for app functionality, I was in an endless loop of being asked to authenticate online. I had to repair by using same authentication logic I used in the Rmd.
3.  Gemini suggested I use some gargle function to clear the cache that did not actually exist, so I was left with many errors.
4.  Generally, if I asked for assistance with one element of the app, the other functionalities would be broken and it was like this for a little.
5.  Need to overly assert that the token should not go on github and what that means in terms of file organization.

That was all, it was moreso loops of me telling it the same thing over and over.
