---
title: "Predicting Prolonged ICU Stay in Sepsis Patients"
author: Yasmine Tani; UID 406579988
format: 
  html:
    embed-resources: true
execute:
  warning: false
  message: false
editor: visual
editor_options: 
  chunk_output_type: inline
---

```{r}
options(readr.show_col_types = FALSE)
```

------------------------------------------------------------------------

------------------------------------------------------------------------

I. Loading Libraries & General Data Inspection

-   looking into CSV variables & columns

-   printing preliminary numbers of patients, unique patients, ICU patients, etc.

------------------------------------------------------------------------

------------------------------------------------------------------------

```{r}
#loading libraries we may need

library(arrow)
library(gtsummary)
library(memuse)
library(pryr)
library(R.utils)
library(tidyverse)
library(stringr)
library(ggplot2)
library(lubridate)
library(readr)
library(randomForest)
library(data.table)
library(glmnet)
```

```{bash}
# we are using icu data so let us again see contents of mimic icu
ls -l ~/mimic/icu
```

```{bash}
# we will also need labs and more from the hopsital data so lets see whats in that folder
# folder and then the head of the admissions csv
ls -l ~/mimic/hosp
```

```{bash}
gzcat ~/mimic/hosp/admissions.csv.gz | head
```

```{bash}
# before i ingest the icu csv i want to look into all these csv's to make sure i know
# where to get the data from and what the variables would look like, etc.
#check column names for admissions.csv.gz [ first few lines]
gzcat ~/mimic/hosp/admissions.csv.gz | head
```

```{bash}
#how many unique patients?
gzcat ~/mimic/hosp/admissions.csv.gz | tail -n +2 | awk -F, '{print $1}' | sort | uniq | wc -l
```

```{bash}
#for our icu data 
#count ICU stays [rows of this]
echo "--- ICU STAYS ---"
gzcat ~/mimic/icu/icustays.csv.gz | tail -n +2 | wc -l

#count unique patients w/ subject_id
echo "--- UNIQUE PATIENTS ---"
gzcat ~/mimic/icu/icustays.csv.gz | tail -n +2 | awk -F, '{print $1}' | sort | uniq | wc -l
```

```{r}
# load the hospital data (where patients and diagnoses live)
patients <- read_csv("mimic/hosp/patients.csv.gz")
admissions <- read_csv("mimic/hosp/admissions.csv.gz")
diagnoses_icd <- read_csv("mimic/hosp/diagnoses_icd.csv.gz")

# load the ICU data (where the stays live)
icu_stays <- read_csv("mimic/icu/icustays.csv.gz")
```

```{r}
library(dplyr)
library(readr)

# load tables
admissions <- read_csv("mimic/hosp/admissions.csv.gz")
icu_stays <- read_csv("mimic/icu/icustays.csv.gz")

# join them together
# use inner_join to keep only those who appear in both tables
icu_admissions <- admissions %>%
  inner_join(icu_stays, by = c("subject_id", "hadm_id"))

# counts
total_admissions <- n_distinct(admissions$hadm_id)
total_icu_admissions <- n_distinct(icu_admissions$hadm_id)
total_icu_stays <- n_distinct(icu_admissions$stay_id)

print(paste("Total Hospital Admissions:", total_admissions))
print(paste("Total Admissions that included an ICU stay:", total_icu_admissions))
print(paste("Total individual ICU stays (stay_id):", total_icu_stays))

```

------------------------------------------------------------------------

------------------------------------------------------------------------

II\. Creation of our Patient Cohort

-   filtering to sepsis patients with their **primary** diagnosis being sepsis

-   utilized a keyword search in the dictionary because we found that this approach yields more patients than merely searching by all the possible ICD codes (there are some specific codes we wont know for additional types of sepsis)

-   filtering to patients in the ICU and only utilizing their first admission to the ICU

-   only keeping patients \>18 \[ adults \]

-   introducing demographic, insurance, lab, comorbidities, vasopressor medications, and vitals data that are relevant to our cohort

-   **thorough** filtration of extreme values & imputation of missing values through necessary means \[ had low missing %'s so we used median imputation\]

-   removal of patients that died within the first 48 hours or model would not be able to distinguish between very health & very sick and would perform very poorly for long stays

------------------------------------------------------------------------

------------------------------------------------------------------------

```{r}
# load the dictionary table
d_icd <- read_csv("mimic/hosp/d_icd_diagnoses.csv.gz")

# identify every code that has sepsis in title
sepsis_keyword_codes <- d_icd %>%
  filter(str_detect(tolower(long_title), "sepsis")) %>%
  pull(icd_code)

# filter for primary diagnosis
primary_sepsis_admissions <- diagnoses_icd %>%
  filter(seq_num == 1, icd_code %in% sepsis_keyword_codes) %>%
  select(hadm_id) %>%
  distinct()

# count
print(paste("Primary Sepsis Patients Admitted:", nrow(primary_sepsis_admissions)))
```

```{r}
# initial join (sepsis in ICU)
sepsis_icu_only <- icu_stays %>% inner_join(primary_sepsis_admissions, by = "hadm_id")
print(paste("Sepsis in ICU Count:", nrow(sepsis_icu_only))) # This is the one line you wanted

# refine to adults and first stays
sepsis_cohort_initial <- sepsis_icu_only %>%
  inner_join(patients, by = "subject_id") %>%
  inner_join(admissions %>% select(hadm_id, race, insurance, hospital_expire_flag), by = "hadm_id") %>%
  filter(!is.na(intime), !is.na(outtime)) %>%
  mutate(age_at_admit = anchor_age + (year(intime) - anchor_year)) %>%
  filter(age_at_admit >= 18, los > 0) %>%
  group_by(subject_id) %>%
  arrange(intime) %>%
  slice(1) %>%
  ungroup()

print(paste("Adults / 1st Stay Count:", nrow(sepsis_cohort_initial)))

# apply the 48-hour survival filter
sepsis_cohort_final <- sepsis_cohort_initial %>%
  filter(!(hospital_expire_flag == 1 & los < 2))

print(paste("Final Cohort Count (Survival >48h):", nrow(sepsis_cohort_final)))

# update data frame for the plot to include all 4 steps
cleaning_steps <- data.frame(
  Step = c("Hospital Sepsis", "ICU Sepsis", "Adults / 1st Stay", "Survival >48h"),
  Count = c(nrow(primary_sepsis_admissions), nrow(sepsis_icu_only), nrow(sepsis_cohort_initial), nrow(sepsis_cohort_final))
)
```

```{r}
#fix the order so it doesn't alphabetize
cleaning_steps$Step <- factor(cleaning_steps$Step, levels = cleaning_steps$Step)

ggplot(cleaning_steps, aes(x = Step, y = Count, fill = Step)) +
  geom_bar(stat = "identity", show.legend = FALSE, width = 0.6) +
  geom_text(aes(label = Count), vjust = -0.5, fontface = "bold") +
  theme_minimal() +
  scale_fill_brewer(palette = "Blues", direction = -1) +
  labs(
    title = "Cohort Selection Attrition",
    x = "",
    y = "Patient Count"
  ) +
  theme(axis.text.x = element_text(angle = 20, vjust = 0.8))
```

```{r}
# calculate average length of stay (ICU)
avg_los <- mean(sepsis_cohort_final$los, na.rm = TRUE)
med_los <- median(sepsis_cohort_final$los, na.rm = TRUE)

# create the 14-day binary target: OUR OUTCOME
sepsis_cohort_final <- sepsis_cohort_final %>%
  mutate(prolonged_stay_14 = if_else(los > 14, 1, 0))

#balance of our labels
table(sepsis_cohort_final$prolonged_stay_14)
```
