---
title: "Predicting Prolonged ICU Stay in Sepsis Patients"
author: Yasmine Tani; UID 406579988
format: 
  html:
    embed-resources: true
execute:
  warning: false
  message: false
editor: visual
editor_options: 
  chunk_output_type: console
---

```{r}
options(readr.show_col_types = FALSE)
```

------------------------------------------------------------------------

------------------------------------------------------------------------

I. Loading Libraries & General Data Inspection

-   looking into CSV variables & columns

-   printing preliminary numbers of patients, unique patients, ICU patients, etc.

------------------------------------------------------------------------

------------------------------------------------------------------------

```{r}
#loading libraries we may need

library(arrow)
library(gtsummary)
library(memuse)
library(pryr)
library(R.utils)
library(tidyverse)
library(stringr)
library(ggplot2)
library(lubridate)
library(readr)
library(randomForest)
library(data.table)
library(glmnet)
```

```{bash}
# we are using icu data so let us again see contents of mimic icu
ls -l ~/mimic/icu
```

```{bash}
# we will also need labs and more from the hopsital data so lets see whats in that folder
# folder and then the head of the admissions csv
ls -l ~/mimic/hosp
```

```{bash}
gzcat ~/mimic/hosp/admissions.csv.gz | head
```

```{bash}
# before i ingest the icu csv i want to look into all these csv's to make sure i know
# where to get the data from and what the variables would look like, etc.
#check column names for admissions.csv.gz [ first few lines]
gzcat ~/mimic/hosp/admissions.csv.gz | head
```

```{bash}
#how many unique patients?
gzcat ~/mimic/hosp/admissions.csv.gz | tail -n +2 | awk -F, '{print $1}' | sort | uniq | wc -l
```

```{bash}
#for our icu data 
#count ICU stays [rows of this]
echo "--- ICU STAYS ---"
gzcat ~/mimic/icu/icustays.csv.gz | tail -n +2 | wc -l

#count unique patients w/ subject_id
echo "--- UNIQUE PATIENTS ---"
gzcat ~/mimic/icu/icustays.csv.gz | tail -n +2 | awk -F, '{print $1}' | sort | uniq | wc -l
```

```{r}
# load the hospital data (where patients and diagnoses live)
patients <- read_csv("mimic/hosp/patients.csv.gz")
admissions <- read_csv("mimic/hosp/admissions.csv.gz")
diagnoses_icd <- read_csv("mimic/hosp/diagnoses_icd.csv.gz")

# load the ICU data (where the stays live)
icu_stays <- read_csv("mimic/icu/icustays.csv.gz")
```

```{r}
library(dplyr)
library(readr)

# load tables
admissions <- read_csv("mimic/hosp/admissions.csv.gz")
icu_stays <- read_csv("mimic/icu/icustays.csv.gz")

# join them together
# use inner_join to keep only those who appear in both tables
icu_admissions <- admissions %>%
  inner_join(icu_stays, by = c("subject_id", "hadm_id"))

# counts
total_admissions <- n_distinct(admissions$hadm_id)
total_icu_admissions <- n_distinct(icu_admissions$hadm_id)
total_icu_stays <- n_distinct(icu_admissions$stay_id)

print(paste("Total Hospital Admissions:", total_admissions))
print(paste("Total Admissions that included an ICU stay:", total_icu_admissions))
print(paste("Total individual ICU stays (stay_id):", total_icu_stays))

```

------------------------------------------------------------------------

------------------------------------------------------------------------

II\. Creation of our Patient Cohort

-   filtering to sepsis patients with their **primary** diagnosis being sepsis

-   utilized a keyword search in the dictionary because we found that this approach yields more patients than merely searching by all the possible ICD codes (there are some specific codes we wont know for additional types of sepsis)

-   filtering to patients in the ICU and only utilizing their first admission to the ICU

-   only keeping patients \>18 \[ adults \]

-   introducing demographic, insurance, lab, comorbidities, vasopressor medications, and vitals data that are relevant to our cohort

-   **thorough** filtration of extreme values & imputation of missing values through necessary means \[ had low missing %'s so we used median imputation\]

-   removal of patients that died within the first 48 hours or model would not be able to distinguish between very health & very sick and would perform very poorly for long stays

------------------------------------------------------------------------

------------------------------------------------------------------------

```{r}
# load the dictionary table
d_icd <- read_csv("mimic/hosp/d_icd_diagnoses.csv.gz")

# identify every code that has sepsis in title
sepsis_keyword_codes <- d_icd %>%
  filter(str_detect(tolower(long_title), "sepsis")) %>%
  pull(icd_code)

# filter for primary diagnosis
primary_sepsis_admissions <- diagnoses_icd %>%
  filter(seq_num == 1, icd_code %in% sepsis_keyword_codes) %>%
  select(hadm_id) %>%
  distinct()

# count
print(paste("Primary Sepsis Patients Admitted:", nrow(primary_sepsis_admissions)))
```

```{r}
# initial join (sepsis in ICU)
sepsis_icu_only <- icu_stays %>% inner_join(primary_sepsis_admissions, by = "hadm_id")
print(paste("Sepsis in ICU Count:", nrow(sepsis_icu_only))) # This is the one line you wanted

# refine to adults and first stays
sepsis_cohort_initial <- sepsis_icu_only %>%
  inner_join(patients, by = "subject_id") %>%
  inner_join(admissions %>% select(hadm_id, race, insurance, hospital_expire_flag), by = "hadm_id") %>%
  filter(!is.na(intime), !is.na(outtime)) %>%
  mutate(age_at_admit = anchor_age + (year(intime) - anchor_year)) %>%
  filter(age_at_admit >= 18, los > 0) %>%
  group_by(subject_id) %>%
  arrange(intime) %>%
  slice(1) %>%
  ungroup()

print(paste("Adults / 1st Stay Count:", nrow(sepsis_cohort_initial)))

# apply the 48-hour survival filter
sepsis_cohort_final <- sepsis_cohort_initial %>%
  filter(!(hospital_expire_flag == 1 & los < 2))

print(paste("Final Cohort Count (Survival >48h):", nrow(sepsis_cohort_final)))

# update data frame for the plot to include all 4 steps
cleaning_steps <- data.frame(
  Step = c("Hospital Sepsis", "ICU Sepsis", "Adults / 1st Stay", "Survival >48h"),
  Count = c(nrow(primary_sepsis_admissions), nrow(sepsis_icu_only), nrow(sepsis_cohort_initial), nrow(sepsis_cohort_final))
)
```

```{r}
#fix the order so it doesn't alphabetize
cleaning_steps$Step <- factor(cleaning_steps$Step, levels = cleaning_steps$Step)

ggplot(cleaning_steps, aes(x = Step, y = Count, fill = Step)) +
  geom_bar(stat = "identity", show.legend = FALSE, width = 0.6) +
  geom_text(aes(label = Count), vjust = -0.5, fontface = "bold") +
  theme_minimal() +
  scale_fill_brewer(palette = "Blues", direction = -1) +
  labs(
    title = "Cohort Selection Attrition",
    x = "",
    y = "Patient Count"
  ) +
  theme(axis.text.x = element_text(angle = 20, vjust = 0.8))
```

```{r}
# calculate average length of stay (ICU)
avg_los <- mean(sepsis_cohort_final$los, na.rm = TRUE)
med_los <- median(sepsis_cohort_final$los, na.rm = TRUE)

# create the 14-day binary target: OUR OUTCOME
sepsis_cohort_final <- sepsis_cohort_final %>%
  mutate(prolonged_stay_14 = if_else(los > 14, 1, 0))

#balance of our labels
table(sepsis_cohort_final$prolonged_stay_14)
```

```{r}
# join with admissions 
sepsis_cohort_final <- sepsis_cohort_final %>%
  mutate(
    # categroize race to reduce noise & maxmiize data use
    race_group = case_when(
      str_detect(race, "WHITE") ~ "WHITE",
      str_detect(race, "BLACK") ~ "BLACK",
      str_detect(race, "HISPANIC") ~ "HISPANIC",
      str_detect(race, "ASIAN") ~ "ASIAN",
      TRUE ~ "OTHER/UNKNOWN"
    ),
    # ensure Insurance is a factor
    insurance = as.factor(insurance)
  )

# breakdown of these features
print("Insurance Breakdown:")
table(sepsis_cohort_final$insurance)

print("Cleaned Race Breakdown:")
table(sepsis_cohort_final$race_group)
```

```{r}
# age distribution
ggplot(sepsis_cohort_final, aes(x = age_at_admit)) + 
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  theme_minimal() + 
  labs(title = "Age Distribution of Sepsis Cohort", x = "Age", y = "Count")

# gender Breakdown
ggplot(sepsis_cohort_final, aes(x = gender, fill = gender)) + 
  geom_bar() + 
  theme_minimal() + 
  labs(title = "Gender Distribution", x = "Gender", y = "Count")
```

```{r}
# lab IDs
lab_id_map <- list(
  lactate = c(50813, 52442, 53154),
  creatinine = c(50912, 52024, 52546),
  bilirubin = c(50885, 53089),
  wbc = c(51301, 51755, 51756),
  platelets = c(51265, 53189)
)

# filter for the first 24 hours of the ICU stay
labs_features_raw <- open_dataset("labevents_pq") %>%
  filter(hadm_id %in% sepsis_cohort_final$hadm_id) %>%
  filter(itemid %in% unlist(lab_id_map)) %>%
  collect() %>%
  inner_join(sepsis_cohort_final %>% select(hadm_id, intime), by = "hadm_id") %>%
  filter(charttime >= intime, charttime <= intime + hours(24)) %>%
  mutate(lab_name = case_when(
    itemid %in% lab_id_map$lactate ~ "max_lactate",
    itemid %in% lab_id_map$creatinine ~ "max_creatinine",
    itemid %in% lab_id_map$bilirubin ~ "max_bilirubin",
    itemid %in% lab_id_map$wbc ~ "max_wbc",
    itemid %in% lab_id_map$platelets ~ "min_platelets"
  )) %>%
  # biological sanitycCheck: remove extreme outliers/errors
  filter(
    !(lab_name == "max_lactate" & valuenum > 30),
    !(lab_name == "max_wbc" & valuenum > 200),
    !(lab_name == "min_platelets" & valuenum > 2000)
  )

# summarize to get the "worst" value per patient
labs_summarized <- labs_features_raw %>%
  group_by(hadm_id, lab_name) %>%
  summarize(
    val = if (all(is.na(valuenum))) {
      NA_real_  # If all values are NA, return a proper NA instead of -Inf
    } else if (str_detect(first(lab_name), "min")) {
      min(valuenum, na.rm = TRUE)
    } else {
      max(valuenum, na.rm = TRUE)
    },
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = lab_name, values_from = val)

print("Labs Summarized Preview:")
labs_summarized %>% select(1:min(5, ncol(.))) %>% head()
```

```{r}
# join labs to the master sepsis cohort
sepsis_cohort_final <- sepsis_cohort_final %>%
  left_join(labs_summarized, by = "hadm_id")

# check missingness before imputation
print("Missing counts BEFORE imputation:")
print(colSums(is.na(sepsis_cohort_final %>% select(starts_with("max_"), starts_with("min_")))))

# median imputation
sepsis_cohort_final <- sepsis_cohort_final %>%
  mutate(across(c(max_lactate, max_creatinine, max_bilirubin, max_wbc, min_platelets), 
                ~ if_else(is.na(.), median(., na.rm = TRUE), .)))

# final verification
print("Missing counts AFTER imputation:")
print(colSums(is.na(sepsis_cohort_final %>% select(starts_with("max_"), starts_with("min_")))))
```

```{r}
ggplot(sepsis_cohort_final, aes(x = max_creatinine, fill = as.factor(prolonged_stay_14))) +
  geom_density(alpha = 0.5) +
  scale_x_log10() + 
  theme_minimal() +
  scale_fill_manual(values = c("skyblue", "orange"), name = "Stay > 14 Days") +
  labs(
    title = "Distribution of Max Creatinine by Stay Length",
    subtitle = "Sepsis patients with stays >14 days tend to have higher peak creatinine levels.",
    x = "Max Creatinine (mg/dL) - Log Scale",
    y = "Density"
  )
```

```{r}
# IDs
vital_ids <- c(220045, 220210, 220181, 220052, 223761, 223762, 220179, 220050)
# join
vitals_cleaned <- open_dataset("chartevents_pq") %>%
  filter(stay_id %in% sepsis_cohort_final$stay_id) %>% # filter before collect!
  filter(itemid %in% vital_ids) %>%
  collect() %>%
  select(-any_of("hadm_id")) %>% 
  mutate(stay_id = as.numeric(stay_id)) %>%
  inner_join(
    sepsis_cohort_final %>% 
      mutate(stay_id = as.numeric(stay_id)) %>% 
      select(stay_id, hadm_id, intime), 
    by = "stay_id"
  ) %>%
  filter(charttime >= intime, charttime <= intime + hours(24))

# summarize: one median value per patient, per vital sign
vitals_summarized <- vitals_cleaned %>%
  mutate(vital_name = case_when(
    itemid == 220045 ~ "heart_rate",
    itemid == 220210 ~ "resp_rate",
    itemid %in% c(220181, 220052) ~ "mean_bp",
    itemid %in% c(220179, 220050) ~ "sys_bp",    # New Systolic BP
    itemid %in% c(223761, 223762) ~ "temp_f"
  )) %>%
  # convert Celsius to Fahrenheit if itemid is 223762
mutate(valuenum = if_else(itemid == 223762, (valuenum * 9/5) + 32, valuenum)) %>%  
  
# biological sanity check: remove extreme outliers/errors
  filter(
    !(vital_name == "heart_rate" & (valuenum < 0 | valuenum > 300)),
    !(vital_name == "resp_rate" & (valuenum < 0 | valuenum > 100)),
    !(vital_name == "mean_bp" & (valuenum < 0 | valuenum > 300)),
    !(vital_name == "sys_bp" & (valuenum < 0 | valuenum > 300)),  # Added Sys BP
    !(vital_name == "temp_f" & (valuenum < 70 | valuenum > 115))
  ) %>%
  group_by(hadm_id, vital_name) %>%
  summarize(val = median(valuenum, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = vital_name, values_from = val)

# check result: should see hadm_id and the 4 vital columns
print("Structure of summarized vitals:")
head(vitals_summarized)
```

```{r}
v_cols <- c("heart_rate", "resp_rate", "mean_bp", "sys_bp", "temp_f")

# join the summarized vitals
sepsis_cohort_final <- sepsis_cohort_final %>%
  select(-any_of(v_cols)) %>%  # wipe out old versions if they exist
  left_join(vitals_summarized, by = "hadm_id")

# print missingness stats
missing_report <- sepsis_cohort_final %>%
  summarise(across(any_of(v_cols), 
                   list(
                     count = ~sum(is.na(.)),
                     percent = ~round(sum(is.na(.)) / n() * 100, 2)
                   )))

print("--- Vital Signs Missingness Report (Pre-Imputation) ---")
print(missing_report)
```

```{r}
# median imputation (we have low missing % values, teammate confirmed with TAs)
sepsis_cohort_final <- sepsis_cohort_final %>%
  mutate(across(any_of(v_cols), 
                ~ if_else(is.na(.), median(., na.rm = TRUE), .)))

# final verification - should be 0 now
print("--- Final Missing Count Check (Vitals) ---")
print(colSums(is.na(sepsis_cohort_final %>% select(any_of(v_cols)))))
```

```{r}
# grab diagnoses and filter for our sepsis crew
diagnoses <- read_csv("~/mimic/hosp/diagnoses_icd.csv.gz") %>%
  filter(hadm_id %in% sepsis_cohort_final$hadm_id)

# flag the comorbidities using regex (checking for both ICD9/10)
comorbidities <- diagnoses %>%
  group_by(hadm_id) %>%
  summarize(
    hypertension = max(if_else(str_detect(icd_code, "^I1[0-5]|^40[1-5]"), 1, 0)),
    diabetes = max(if_else(str_detect(icd_code, "^E0[8-9]|^E1[0-3]|^250"), 1, 0)),
    ckd = max(if_else(str_detect(icd_code, "^N18|^585"), 1, 0)),
    cvd = max(if_else(str_detect(icd_code, "^I2[0-5]|^I50|^41[0-4]|^428"), 1, 0)),
    liver_disease = max(if_else(str_detect(icd_code, "^B18|^K7[0-6]|^571"), 1, 0)),
    immunosuppression = max(if_else(str_detect(icd_code, "^D8[0-9]|^279|^V581|^Z511"), 1, 0))
  ) %>%
  ungroup()

# join to the master table and clean up
# use sepsis_cohort_final here so we keep our vitals and labs!
sepsis_cohort_final <- sepsis_cohort_final %>%
  select(-any_of(colnames(comorbidities)[-1])) %>% # avoid duplicate col drama
  left_join(comorbidities, by = "hadm_id") %>%
  # if no code code, assume they don't have the condition (set to 0)
  mutate(across(c(hypertension, diabetes, ckd, cvd, liver_disease, immunosuppression),
                ~ if_else(is.na(.), 0, .))) %>%
  # make them factors so the Random Forest treats them as categories, not numbers
  mutate(across(c(hypertension, diabetes, ckd, cvd, liver_disease, immunosuppression), as.factor))

# distribution
summary(sepsis_cohort_final %>% select(hypertension, diabetes, ckd))
```

```{r}
# prevalence of comorbidities
comorb_plot <- sepsis_cohort_final %>%
  select(hypertension, diabetes, ckd, cvd) %>%
  mutate(across(everything(), ~ as.numeric(as.character(.)))) %>%
  summarise(across(everything(), mean)) %>%
  pivot_longer(everything(), names_to = "Condition", values_to = "Prevalence")

ggplot(comorb_plot, aes(x = reorder(Condition, Prevalence), y = Prevalence)) +
  geom_bar(stat = "identity", fill = "darkseagreen") +
  scale_y_continuous(labels = scales::percent) +
  coord_flip() +
  theme_minimal() +
  labs(title = "Prevalence of Comorbidities in Sepsis Cohort", x = "", y = "Percentage of Patients")
```

```{r}
# load prescriptions and filter for our cohort
# using prescriptions for vasopressors is good, though some are in inputevents
prescriptions_raw <- read_csv("~/mimic/hosp/prescriptions.csv.gz") %>%
  filter(hadm_id %in% sepsis_cohort_final$hadm_id)

# flag vasopressors in the first 24 hours
vaso_flags <- prescriptions_raw %>%
  left_join(sepsis_cohort_final %>% select(hadm_id, intime), by = "hadm_id") %>%
  # filter: was the drug started within 24 hours of ICU admission?
  filter(starttime >= intime & starttime <= (intime + hours(24))) %>%
  mutate(drug = tolower(drug)) %>% 
  group_by(hadm_id) %>%
  summarize(
    vasopressors = max(if_else(str_detect(drug, "norepinephrine|epinephrine|phenylephrine|vasopressin|dopamine"), 1, 0)),
    .groups = "drop"
  )

# join back to the master table
sepsis_cohort_final <- sepsis_cohort_final %>%
  # remove old version if it exists
  select(-any_of("vasopressors")) %>% 
  left_join(vaso_flags, by = "hadm_id") %>%
  # if a patient wasn't in the vaso_flags list, they didn't get pressors (set to 0)
  mutate(vasopressors = if_else(is.na(vasopressors), 0, as.numeric(vasopressors))) %>%
  # make it a factor for the model
  mutate(vasopressors = as.factor(vasopressors))

# final Count Check
print("Distribution of Vasopressor use in first 24h:")
table(sepsis_cohort_final$vasopressors)
```

```{r}
# check for any remaining NAs in your potential predictors
colSums(is.na(sepsis_cohort_final %>% 
                select(hospital_expire_flag, age_at_admit, gender, race_group, 
                       starts_with("max_"), starts_with("min_"),
                       heart_rate, mean_bp, sys_bp, resp_rate, temp_f,
                       hypertension, diabetes, ckd, cvd, vasopressors)))
```

```{r}
# checking for weird outliers or extreme values
sepsis_cohort_final %>%
  select(age_at_admit, heart_rate, resp_rate, mean_bp, sys_bp, temp_f, 
         max_lactate, max_creatinine, max_bilirubin) %>%
  summary()
```

```{r}
sepsis_cohort_final <- sepsis_cohort_final %>%
  mutate(
    # fix 0 respiratory rates using the median (if it were a prolonged 0 average then the patient would be dead...)
    resp_rate = if_else(resp_rate <= 0, median(resp_rate[resp_rate > 0], na.rm = TRUE), resp_rate),
    # fix the 0 mean_bp using the median
    mean_bp = if_else(mean_bp <= 0, median(mean_bp[mean_bp > 0], na.rm = TRUE), mean_bp),
    # ensure our stay flag is a factor for the model
    prolonged_stay_14 = as.factor(prolonged_stay_14)
  )

# verify 0 is gone
summary(sepsis_cohort_final$resp_rate)
```

```{r}
# list all column names to make sure everything joined correctly
colnames(sepsis_cohort_final)
```

------------------------------------------------------------------------

------------------------------------------------------------------------
